---
title: "Supplementary Material - Comparing Microbial Diversity and Co-Occurrence Networks Across Five Major Body Sites (HMP 16S)"
author: "Kexin Cui"
date: today
format:
  html:
    toc: true
    number-sections: true
bibliography: ../../../assets/dataanalysis-references.bib
csl: ../../../assets/vancouver-author-date.csl
---

```{r, echo=FALSE, message=FALSE}
library(here)
library(knitr)
library(dplyr)
```

# Overview

This supplement documents (a) the full set of files and the order in which
they must be run to reproduce every figure and table in the main manuscript,
(b) additional methodological detail, and (c) extra results that did not fit
into the main text (pairwise alpha diversity comparisons, detailed network
summaries).

# Code and file information

The project has three code stages. Run them in the following order:

1. `code/processing-code/processingfile-v1.qmd` — downloads the HMP V3–V5
   data with `HMP16SData::V35()`, removes zero-count and rare taxa, and
   writes `data/processed-data/ps_filt.rds` and
   `data/processed-data/ps_rel.rds`.
2. `code/eda-code/eda.qmd` — produces the exploratory figures
   (`seq_depth.png`, `top_phyla.png`, `alpha_diversity_5sites.png`,
   `shannon_boxplot.png`, `pcoa_preview.png`) and the summary table
   `results/tables/exploratory_summary_by_subsite.rds`.
3. `code/analysis-code/statistical-analysis.R` — runs the formal hypothesis
   tests (Kruskal-Wallis, pairwise Wilcoxon, PERMANOVA, BETADISPER),
   builds per-site co-occurrence networks, and trains the supervised
   body-site classifier with cross-validation, model comparison and
   bootstrap-based uncertainty. Outputs include all tables
   (`alpha_kruskal.rds`, `alpha_pairwise.rds`, `permanova.rds`,
   `network_stats.rds`, `ml_cv_metrics.rds`, `ml_test_metrics.rds`,
   `ml_test_uncertainty.rds`, `ml_best_params.rds`,
   `ml_confusion_matrix.rds`, `ml_variable_importance.rds`) and figures
   (`pcoa_ordination_5sites.png`, `hierarchical_clustering_5sites.png`,
   `network_statistics_comparison.png`, `ml_cv_comparison.png`,
   `ml_confusion_matrix.png`, `ml_roc_curves.png`,
   `ml_variable_importance.png`).

The manuscript Quarto file `products/manuscript/Manuscript.qmd` and this
supplement pull these results in via `readRDS()` and
`knitr::include_graphics()`.

{{< pagebreak >}}

# Additional method details

## Stratified subsampling

To make the comparison across the five major body sites balanced, we sample
up to 200 samples per site using base R's `sample()` with `set.seed(123)`.
Sites with fewer than 200 samples are included in full. This procedure gives
the same set of samples on every run.

## Filtering and rarefaction

Taxa are retained only if they are present in at least 10 samples and have a
total count of ≥ 50 reads. Samples are rarefied to the 5th percentile of the
distribution of sample sums, using `phyloseq::rarefy_even_depth()` with
`rngseed = 123`. This threshold keeps 95% of samples while substantially
reducing variation in sequencing depth.

## Distance choice

We use Bray-Curtis as the primary distance because it is abundance-aware
and is the default distance in most HMP-era papers. Jaccard (presence/
absence) is used as a robustness check. BETADISPER tests whether the
PERMANOVA effect is confounded with unequal dispersion among groups; both
tests are reported.

## Network construction

Correlation networks are built per body site from the prevalence-filtered
OTU table (≥ 10% prevalence) using Spearman correlations via `Hmisc::rcorr`.
Edges are retained when |r| ≥ 0.5 and p < 0.01 (no multiple-testing
correction is applied inside the network — this is a pragmatic threshold,
not a formal significance test). Louvain community detection
(`igraph::cluster_louvain`) is used to compute modularity and communities.

## Supervised classification (H4)

The H4 pipeline showcases several data-science topics from the course in
one place: model fitting and tuning, cross-validation, model comparison,
regularization, ensemble methods, calibration via ROC curves, variable
importance, and uncertainty quantification.

- *Features.* The 300 most-abundant taxa, log(1 + x)-transformed and
  z-scored. Constant-zero predictors are removed with `step_zv`.
- *Resampling.* A 75/25 stratified train/test split, with 5-fold
  cross-validation on the training set, also stratified by body site.
- *Candidate models.* Five workflows:
    1. `parsnip::null_model` — baseline that always predicts the modal
       class.
    2. `multinom_reg` (glmnet) — elastic-net regularized multinomial
       logistic regression; `penalty` swept on a log-grid from 1e-4 to 1
       and `mixture` swept from 0 (ridge) to 1 (lasso).
    3. `decision_tree` (rpart) — single CART; `cost_complexity` from 1e-4
       to 1e-1, `tree_depth` from 3 to 15, and `min_n` from 2 to 20.
    4. `rand_forest` (ranger, 500 trees, `importance = "permutation"`) —
       `mtry` swept on \{5, 35, 70, 100\} and `min_n` on \{2, 8, 14, 20\}.
    5. `boost_tree` (xgboost) — `trees` \{100, 800\}, `tree_depth`
       \{3, 9\}, `learn_rate` swept on a log-grid from 1e-2 to ~3e-1, and
       `min_n` \{2, 20\}. Skipped automatically when `xgboost` is not
       installed.
- *Selection.* For each model we pick the hyperparameter setting that
  maximizes mean macro-AUC across the 5 folds, then refit on the full
  training set and evaluate on the held-out test set.
- *Uncertainty.* Test-set accuracy and macro-AUC are bootstrapped on the
  predictions: 1000 resamples (with replacement) of the per-row
  prediction tibble, and the 2.5%/97.5% quantiles serve as the 95%
  confidence interval.
- *Variable importance.* For the winning model, we extract the top-20
  features by `vip::vi`, which uses the engine-native importance
  (permutation importance for ranger, gain for xgboost, and absolute
  coefficient size for glmnet).

{{< pagebreak >}}

# Additional results

## Pairwise alpha diversity comparisons

Table @tbl-alpha-pairwise lists the Benjamini-Hochberg-adjusted
pairwise Wilcoxon p-values for every pair of body sites, for each of the
four alpha-diversity indices. These follow up the overall Kruskal-Wallis
tests reported in the main text.

```{r, echo=FALSE}
#| label: tbl-alpha-pairwise
#| tbl-cap: "Pairwise Wilcoxon (BH-adjusted) p-values for alpha diversity differences across body sites."
pw_path <- here::here("results", "tables", "alpha_pairwise.rds")
if (file.exists(pw_path)) {
  pw <- readRDS(pw_path)
  knitr::kable(pw, digits = 4)
} else {
  cat("Run statistical-analysis.R to produce this table.")
}
```

## Full network statistics

Table @tbl-network-full provides the full per-site network statistics
(reproduced here from the main text for convenience, together with the
number of positive vs. negative edges that are not shown in the main
figure).

```{r, echo=FALSE}
#| label: tbl-network-full
#| tbl-cap: "Per-site co-occurrence network summary."
net_path <- here::here("results", "tables", "network_stats.rds")
if (file.exists(net_path)) {
  ns <- readRDS(net_path)
  knitr::kable(ns, digits = 3)
} else {
  cat("Run statistical-analysis.R to produce this table.")
}
```

## Individual network visualizations

The networks themselves are saved as an `igraph` list in
`results/output/networks.rds`. Each network can be loaded individually and
visualized with `ggraph`; below is a placeholder for a figure panel showing
one representative network per body site.

```{r, echo=FALSE}
#| label: fig-network-oral
#| fig-cap: "Representative co-occurrence network visualization (replace with a saved network PNG)."
fig_path <- here::here("results", "figures",
                       "network_statistics_comparison.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

## Cross-validation comparison and tuned hyperparameters

Table @tbl-ml-cv-supp gives the full cross-validation summary across the
five candidate models, and Table @tbl-ml-best-params lists the
hyperparameters selected by `select_best(metric = "roc_auc")` for each
non-null model.

```{r, echo=FALSE}
#| label: tbl-ml-cv-supp
#| tbl-cap: "Cross-validation accuracy and macro-AUC for every candidate model (mean +/- SD across folds)."
p <- here::here("results", "tables", "ml_cv_metrics.rds")
if (file.exists(p)) knitr::kable(readRDS(p), digits = 3)
```

```{r, echo=FALSE}
#| label: tbl-ml-best-params
#| tbl-cap: "Tuned hyperparameter values selected on cross-validation macro-AUC."
p <- here::here("results", "tables", "ml_best_params.rds")
if (file.exists(p)) knitr::kable(readRDS(p))
```

## Test-set ROC curves and bootstrap uncertainty

Figure @fig-ml-roc-supp shows the per-class ROC curves for the winning
model, and Table @tbl-ml-uncert-supp reports bootstrap 95% confidence
intervals on test accuracy and macro-AUC (B = 1000 resamples). The CIs are
narrow on this dataset, which reflects the large held-out test sample and
the strong signal in the features.

```{r, echo=FALSE}
#| label: fig-ml-roc-supp
#| fig-cap: "Test-set one-vs-rest ROC curves for the winning classifier."
fig_path <- here::here("results", "figures", "ml_roc_curves.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

```{r, echo=FALSE}
#| label: tbl-ml-uncert-supp
#| tbl-cap: "Bootstrap 95% CIs on test-set accuracy and macro-AUC (winning model)."
p <- here::here("results", "tables", "ml_test_uncertainty.rds")
if (file.exists(p)) knitr::kable(readRDS(p), digits = 3)
```

## Variable importance

```{r, echo=FALSE}
#| label: fig-ml-vip-supp
#| fig-cap: "Top 20 features (taxa) for the winning classifier, ranked by importance."
fig_path <- here::here("results", "figures", "ml_variable_importance.png")
if (file.exists(fig_path)) knitr::include_graphics(fig_path)
```

{{< pagebreak >}}

# Discussion

Results in this supplement support the three main-text conclusions: (1) the
overall Kruskal-Wallis tests are not driven by a single outlier site — most
pairwise contrasts are significant after BH correction; (2) PERMANOVA
effects are robust to switching between Bray-Curtis and Jaccard distances;
and (3) network differences across sites are not artifacts of sample size
alone, because the ranking of sites by modularity and communities is
consistent with the ecological structure reported in the literature
[@mckay2020; @mckay2020a].

{{< pagebreak >}}

# References
